---
title: "Predicting Lake Mixed-Layer Temperature with Multiple Linear Regression"
author: "Sarah Dias"
date: "2026-05-20"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)

# All packages used in this analysis, loaded once
library(terra)      # read / handle the GRIB raster data
library(MASS)       # stepAIC, boxcox
library(car)        # VIF, influencePlot
library(glmnet)     # Lasso
library(lmtest)     # Breusch-Pagan, Durbin-Watson
library(ggplot2)    # plots
library(gridExtra)  # arrange plots in a grid
library(caret)      # grouped cross-validation
```

# 1. Data Preparation

## Data Loading and Pre-processing
```{r}
# Path to the ERA5-Land GRIB extract. Download it from the Copernicus CDS and
# place it beside this .Rmd (see the README for the link and region/dates).
grib_file <- "7120240a16eacaafc6ddb2d80c2b6594.grib"

# Stop early with a clear message if the data file is missing.
if (!file.exists(grib_file)) {
  stop("GRIB file not found. Place it in the same folder as this R Markdown file.")
}
```

## Load the GRIB file with terra
```{r}
r <- rast(grib_file)
nlyr(r)
names(r)[1:10]
```
The file holds five variables per timestamp: 2 m temperature, total cloud cover, lake bottom temperature, lake mixed-layer temperature, and surface solar radiation.

```{r}
terra::describe(grib_file)[31:80]
```
The layers repeat in this fixed order through layer 865:
Layer 1: 2 metre temperature [C]
Layer 2: Total cloud cover (0 - 1) [-]
Layer 3: var10 of table 228 of center ECMWF (lake bottom temperature)
Layer 4: var8 of table 228 of center ECMWF (lake mix-layer temperature)
Layer 5: Surface solar radiation [W*s/m^2]

## Split the stack into one raster per variable
Based on the GRIB metadata:
var10 of table 228 of center ECMWF = lake bottom temperature
var8 of table 228 of center ECMWF = lake mix-layer temperature
Source: https://gdex.ucar.edu/datasets/d633001/metadata/grib.html

```{r}
# Every 5th layer belongs to the same variable (order is fixed above).
t2m              <- r[[seq(1, nlyr(r), by = 5)]]  # 2 m temperature
cloud            <- r[[seq(2, nlyr(r), by = 5)]]  # total cloud cover
lake_bottom_temp <- r[[seq(3, nlyr(r), by = 5)]]  # lake bottom temperature
lake_mix_temp    <- r[[seq(4, nlyr(r), by = 5)]]  # lake mixed-layer temperature
solar            <- r[[seq(5, nlyr(r), by = 5)]]  # surface solar radiation
```

## Confirm each variable has 173 layers
```{r}
nlyr(t2m)
nlyr(cloud)
nlyr(lake_bottom_temp)
nlyr(lake_mix_temp)
nlyr(solar)
```

## Inspect the timestamps
```{r}
times <- time(t2m)
times[1:5]
times[168:173]
```
Each variable has 173 timestamps: two winter dates per year (January 1 and December 1), from 1940 through January 2026.

## Spatially average each layer over the region
Each layer is a map over the Lake Superior grid, not a single value. To fit standard multiple linear regression with one observation per timestamp, average every grid cell in each layer to a single regional mean.

```{r}
# global() computes a summary statistic (here, the mean) over the whole grid.
mean_t2m    <- global(t2m,              "mean", na.rm = TRUE)[, 1]
mean_cloud  <- global(cloud,            "mean", na.rm = TRUE)[, 1]
mean_bottom <- global(lake_bottom_temp, "mean", na.rm = TRUE)[, 1]
mean_mix    <- global(lake_mix_temp,    "mean", na.rm = TRUE)[, 1]
mean_solar  <- global(solar,            "mean", na.rm = TRUE)[, 1]
```

## Build the regression table
One row per timestamp; columns are the response and the four predictors.
```{r}
model_data <- data.frame(
  t2m              = mean_t2m,
  cloud            = mean_cloud,
  lake_bottom_temp = mean_bottom,
  lake_mix_temp    = mean_mix,
  solar            = mean_solar
)
head(model_data)
summary(model_data)
```

## Check for missing values
```{r}
colSums(is.na(model_data))
```
There are no missing values.

Summary of preprocessing: each raster layer was averaged across the region to one regional mean per variable per timestamp, producing a time-indexed regression table with lake mixed-layer temperature as the response and 2 m temperature, cloud cover, lake bottom temperature, and surface solar radiation as predictors.

# 2. Exploratory Data Analysis

## First layer of each variable
```{r}
par(mfrow = c(2, 3))
plot(t2m[[1]],              main = "2m Temperature - Layer 1")
plot(cloud[[1]],            main = "Cloud Cover - Layer 1")
plot(lake_bottom_temp[[1]], main = "Bottom Temp - Layer 1")
plot(lake_mix_temp[[1]],    main = "Mix-Layer Temp - Layer 1")
plot(solar[[1]],            main = "Solar Radiation - Layer 1")
par(mfrow = c(1, 1))
```

## Pairwise scatterplots
```{r}
pairs(model_data[, c("lake_mix_temp", "t2m", "cloud", "lake_bottom_temp", "solar")])
```

## Correlation matrix
```{r}
cor(model_data[, c("lake_mix_temp", "t2m", "cloud", "lake_bottom_temp", "solar")])
```

# 3. Model Fitting

## Full model (all predictors)
```{r}
full_model <- lm(lake_mix_temp ~ t2m + cloud + lake_bottom_temp + solar,
                 data = model_data)
summary(full_model)
```

## Variable selection

### Stepwise AIC (backward)
```{r}
step_model <- stepAIC(full_model, direction = "backward")
summary(step_model)
```

### Stepwise AIC (forward) — cross-check
```{r}
null_model <- lm(lake_mix_temp ~ 1, data = model_data)

step_forward <- stepAIC(
  null_model,
  scope     = list(lower = ~1,
                   upper = ~ t2m + cloud + lake_bottom_temp + solar),
  direction = "forward"
)
```

### Stepwise BIC
```{r}
n <- nrow(model_data)
step_model_bic <- stepAIC(full_model, direction = "backward", k = log(n))
```

### Lasso
```{r}
X <- as.matrix(model_data[, c("t2m", "cloud", "lake_bottom_temp", "solar")])
y <- model_data$lake_mix_temp

lasso_model <- cv.glmnet(X, y, alpha = 1)  # alpha = 1 is Lasso
plot(lasso_model)
coef(lasso_model, s = "lambda.min")  # coefficients at the best lambda
coef(lasso_model, s = "lambda.1se")  # coefficients at the 1-SE lambda
```

Lasso keeps all four predictors — none are shrunk to zero. `solar` is very close to zero (-2.656e-07), i.e. it contributes little, which matches it being the weakest predictor under AIC/BIC.

**Variable selection results:** stepwise AIC (forward and backward), BIC, and Lasso all retain the four predictors.

## Multicollinearity (VIF)
```{r}
vif(full_model)
```
All VIFs are below 2, so multicollinearity is not a concern.

## Testing for interactions

Physically plausible interactions:

1. `solar × cloud` — cloud cover blocks solar radiation, so solar's effect on lake temperature likely depends on cloudiness. Strongest candidate.
2. `t2m × cloud` — cloud cover has an insulating effect on air temperature, so their combined effect may not be additive.
3. `t2m × solar` — higher air temperature plus more solar radiation could compound warming.

### Full two-way interaction model
```{r}
interaction_model <- lm(lake_mix_temp ~ (t2m + cloud + lake_bottom_temp + solar)^2,
                        data = model_data)
summary(interaction_model)

# Compare against the additive full model.
AIC(full_model, interaction_model)
anova(full_model, interaction_model)
```

`t2m:lake_bottom_temp` (p = 4.56e-07) and `cloud:solar` (p = 0.0453) are significant at the 0.05 level.

### Refined model with the significant interactions
```{r}
refined_model <- lm(lake_mix_temp ~ t2m + cloud + lake_bottom_temp + solar +
                      t2m:lake_bottom_temp +
                      cloud:solar,
                    data = model_data)
summary(refined_model)
AIC(full_model, refined_model, interaction_model)
```

AIC drops well below the full model and adjusted R^2 rises to 0.7926.

## Investigating quadratic terms

### Linear vs. quadratic fits for each predictor
```{r}
p1 <- ggplot(model_data, aes(x = t2m, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x,         aes(color = "Linear"),    se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "t2m vs lake_mix_temp", color = "Fit") +
  theme_minimal()

p2 <- ggplot(model_data, aes(x = lake_bottom_temp, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x,         aes(color = "Linear"),    se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "lake_bottom_temp vs lake_mix_temp", color = "Fit") +
  theme_minimal()

p3 <- ggplot(model_data, aes(x = cloud, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x,         aes(color = "Linear"),    se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "cloud vs lake_mix_temp", color = "Fit") +
  theme_minimal()

p4 <- ggplot(model_data, aes(x = solar, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x,         aes(color = "Linear"),    se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "solar vs lake_mix_temp", color = "Fit") +
  theme_minimal()

grid.arrange(p1, p2, p3, p4, ncol = 2)
```

### Full quadratic model
```{r}
full_model_quad <- lm(lake_mix_temp ~ t2m + I(t2m^2) +
                        cloud + I(cloud^2) +
                        lake_bottom_temp + I(lake_bottom_temp^2) +
                        solar + I(solar^2) +
                        t2m:lake_bottom_temp + cloud:solar, data = model_data)
summary(full_model_quad)
```
Only the quadratic term `I(t2m^2)` is significant at the 0.05 level (p = 0.0012).

### Refined quadratic model
```{r}
refined_model_quad <- lm(lake_mix_temp ~ t2m + I(t2m^2) + cloud +
                           lake_bottom_temp + solar + t2m:lake_bottom_temp + cloud:solar,
                         data = model_data)
summary(refined_model_quad)
AIC(refined_model_quad)
```

## Checking assumptions

### Diagnostic plots
```{r}
par(mfrow = c(2, 2))
plot(refined_model_quad)
par(mfrow = c(1, 1))
```
The diagnostics suggest the normality and constant-variance assumptions are not fully met: the Q-Q plot shows heavy tails and the scale-location plot shows some spread in the residuals. The formal tests below quantify this.

### Shapiro-Wilk normality test
```{r}
shapiro.test(residuals(refined_model_quad))
```

### Breusch-Pagan test (constant variance)
```{r}
bptest(refined_model_quad)
```

### Durbin-Watson test (autocorrelation)
```{r}
# Autocorrelation is worth checking because the data is time-indexed.
dwtest(refined_model_quad)
```

## Trying transformations

### Anomaly (mean-centering) transformation
A common climate-science step: subtract each variable's mean to work in anomalies.
```{r}
model_data$lake_mix_temp_anom    <- model_data$lake_mix_temp    - mean(model_data$lake_mix_temp)
model_data$t2m_anom              <- model_data$t2m              - mean(model_data$t2m)
model_data$lake_bottom_temp_anom <- model_data$lake_bottom_temp - mean(model_data$lake_bottom_temp)
model_data$cloud_anom            <- model_data$cloud            - mean(model_data$cloud)
model_data$solar_anom            <- model_data$solar            - mean(model_data$solar)
```

```{r}
# Note: I(t2m_anom^2) — the square must live INSIDE I(), or the formula's
# ^ operator would drop the quadratic term rather than square the variable.
refined_quad_model_anom <- lm(lake_mix_temp_anom ~ t2m_anom + I(t2m_anom^2) +
                                cloud_anom + lake_bottom_temp_anom + solar_anom +
                                t2m_anom:lake_bottom_temp_anom +
                                cloud_anom:solar_anom,
                              data = model_data)

summary(refined_quad_model_anom)
par(mfrow = c(2, 2))
plot(refined_quad_model_anom)
par(mfrow = c(1, 1))
```
Centering the response and predictors only shifts the intercept in an OLS fit — the fitted values, residuals, R^2, RSE, and every residual-based diagnostic are identical to the untransformed `refined_model_quad`. So this step confirms that a mean/anomaly transformation cannot address the normality or variance issues here; a transformation of the response itself is needed instead (Box-Cox, below).

### Box-Cox transformation
```{r}
bc <- boxcox(refined_model_quad)
optimal_lambda <- bc$x[which.max(bc$y)]
optimal_lambda

# Apply the Box-Cox transformation to the response.
model_data$lake_mix_temp_bc <- (model_data$lake_mix_temp^optimal_lambda - 1) / optimal_lambda

model_boxcox <- lm(lake_mix_temp_bc ~ t2m + I(t2m^2) + cloud +
                     lake_bottom_temp + solar + t2m:lake_bottom_temp + cloud:solar,
                   data = model_data)

summary(model_boxcox)
par(mfrow = c(2, 2))
plot(model_boxcox)
par(mfrow = c(1, 1))
```
The Box-Cox fit does not meaningfully improve the diagnostics and costs interpretability, so the untransformed response is retained.

## Investigating outliers

### Influence plot
```{r}
influencePlot(refined_model_quad)
```

### Inspect the flagged observation
```{r}
# The influence plot flags one observation with the largest Cook's distance.
outlier_index <- 118
model_data[outlier_index, ]
```
Observation 118 is the maximum lake-bottom-temperature reading in the series and is the most influential point, so it is refit without below to check its leverage on the model.

### Final model (influential observation removed)
```{r}
final_model <- lm(lake_mix_temp ~ t2m + I(t2m^2) + cloud +
                    lake_bottom_temp + solar + t2m:lake_bottom_temp + cloud:solar,
                  data = model_data[-outlier_index, ])
summary(final_model)
par(mfrow = c(2, 2))
plot(final_model)
par(mfrow = c(1, 1))
AIC(final_model)
```

# 4. Out-of-Sample Validation

## Prepare the data
```{r}
set.seed(8901)

# Rebuild the time index and derive a year label. Each year contributes two
# consecutive timestamps (Jan 1 and Dec 1), so integer-dividing by 2 groups them.
model_data$time_index <- 1:173
model_data$year       <- 1940 + (model_data$time_index - 1) %/% 2
```

```{r}
# Verify consecutive pairs of timestamps share a year.
head(model_data$year)
```

```{r}
# Drop the same influential observation used for the final model.
cv_data <- model_data[-outlier_index, ]
```

## Grouped 10-fold cross-validation
Folds are built on `year` so the two timestamps from the same year are never split across a train/test boundary — this avoids leakage between paired observations.

```{r}
folds <- groupKFold(cv_data$year, k = 10)
```

```{r}
train_control <- trainControl(index = folds, method = "cv")

cv_model <- train(
  lake_mix_temp ~ t2m + I(t2m^2) + cloud + lake_bottom_temp + solar +
                  t2m:lake_bottom_temp + cloud:solar,
  data      = cv_data,
  method    = "lm",
  trControl = train_control
)

print(cv_model)
```

Note on interpretation: the model structure (which interaction and quadratic terms to keep, and removing the influential observation) was chosen using the full dataset before cross-validating the fixed formula. The cross-validated R^2 is therefore a mild optimistic estimate, since it does not capture the uncertainty in model selection itself.
