---
title: "environmental_stats_final_project"
author: "Sarah Dias"
date: "2026-05-20"
output: pdf_document
---

# 1. Data Preparation

## Data Loading and Pre-processing
```{r}
# The following file is our data:
# 7120240a16eacaafc6ddb2d80c2b6594.grib
# Put this GRIB file in the same folder as this RMarkdown file
grib_file <- "7120240a16eacaafc6ddb2d80c2b6594.grib"
# We can only move forward if we have uploaded our GRIB file:
if (!file.exists(grib_file)) {
  stop("GRIB file wasn't found, it has to be in the same folder as this RMarkdown file.")
}
```

## Use Terra package to unload the grib file
```{r}
library(terra)
r <- rast(grib_file)
nlyr(r)
names(r)[1:10]

```
We have 5 variables: 2m temperature, total cloud cover, undefined, undefined, surface solar radiation.

```{r}
terra::describe("7120240a16eacaafc6ddb2d80c2b6594.grib")[31:80]
```
We have these layers in the following order:
Layer 1: 2 metre temperature [C]
Layer 2: Total cloud cover (0 - 1) [-]
Layer 3: var10 of table 228 of center ECMWF (lake bottom temperature)
Layer 4: var8 of table 228 of center ECMWF (lake mix-layer temperature)
Layer 5: Surface solar radiation [W*s/m^2]

And this gets repeated until layer 865

## Split into layers for each variable
Based on GRIB metadata:
var10 of table 228 of center ECMWF = lake bottom temperature
var8 of table 228 of center ECMWF = lake mix-layer temperature
Source: https://gdex.ucar.edu/datasets/d633001/metadata/grib.html

## Naming the layers
```{r}
# take layers 1, 6, 11, ... (correspondent to 2m temperature layers)
t2m <- r[[seq(1, nlyr(r), by = 5)]]
# take layers 2, 7, 12, ... (correspondent to cloud coverage)
cloud <- r[[seq(2, nlyr(r), by = 5)]]
# take layers 3, 8, 13, ... (correspondent to lake bottom temperature)
lake_bottom_temp <- r[[seq(3, nlyr(r), by = 5)]]
# take layers 4, 9, 14, ... (correspondent to lake mixed temperature)
lake_mix_temp <- r[[seq(4, nlyr(r), by = 5)]]
# take layers 5, 10, 15, ... (correspondent to surface solar radiation)
solar <- r[[seq(5, nlyr(r), by = 5)]]
```

## Double check that there are 173 layers for each variable
```{r}
nlyr(t2m)
nlyr(cloud)
nlyr(lake_bottom_temp)
nlyr(lake_mix_temp)
nlyr(solar)
```
## Inspect each timestamp
```{r}
times <- time(t2m)
times[1:5]
times[168:173]
```
Each variable has 173 timestamps. The extracted data includes two winter-period timestamps per year, January 1 and December 1, from 1940 through January 2026.


## Spatially average each raster layer over the selected region
We convert each map at each time step into one regional mean value. To make one clean dataset for regression, we average each variable over the lake grid area for every step:

The global() function from the terra package computes global statistics (like the mean) for an entire raster object (our grid).
```{r}
mean_t2m <- global(t2m, "mean", na.rm = TRUE)[,1]
mean_cloud <- global(cloud, "mean", na.rm = TRUE)[,1]
mean_bottom <- global(lake_bottom_temp, "mean", na.rm = TRUE)[,1]
mean_mix <- global(lake_mix_temp, "mean", na.rm = TRUE)[,1]
mean_solar <- global(solar, "mean", na.rm = TRUE)[,1]

```


## Create final dataset with mean values
One row is one time stamp
Columns is response variable or predictor variable
```{r}
model_data <- data.frame(
t2m = mean_t2m,
cloud = mean_cloud,
lake_bottom_temp = mean_bottom,
lake_mix_temp = mean_mix,
solar = mean_solar
)
head(model_data)
summary(model_data)
```

## Check for missing values
```{r}
# We need to check our missing values before we remove them:
colSums(is.na(model_data))
```
There are no missing values in our dataset

Summary of steps taking for data loading and preprocessing: Each original layer was a raster map over the selected region, not a single value. To use standard multiple linear regression with one observation per time step, we averaged all grid cells in each layer. This produced one regional mean value per variable per timestamp. The final dataset is a time-indexed regression table where the response variable is lake mixed-layer temperature and the predictors are 2m temperature, cloud cover, lake bottom temperature, and surface solar radiation.

# 2. EDA: Exploratory Data Analysis

## First layer of each variable:
```{r}
par(mfrow = c(2, 3))
plot(t2m[[1]], main = "2m Temperature - Layer 1")
plot(cloud[[1]], main = "Cloud Cover - Layer 1")
plot(lake_bottom_temp[[1]], main = "Bottom Temp - Layer 1")
plot(lake_mix_temp[[1]], main = "Mix-Layer Temp - Layer 1")
plot(solar[[1]], main = "Solar Radiation - Layer 1")
par(mfrow = c(1, 1))
```
## Pairs Plot
```{r}
# Pairwise scatterplots to inspect relationships among variables
pairs(model_data[, c("lake_mix_temp", "t2m", "cloud", "lake_bottom_temp", "solar")])
```

## Correlation matrix:
```{r}
cor(model_data[,c("lake_mix_temp", "t2m", "cloud", "lake_bottom_temp", "solar")])
```


# 3. Model Fitting
## Simple Multiple Linear Regression with all variables (full model)
```{r}
# Fit the full model first
full_model <- lm(lake_mix_temp ~ t2m + cloud + lake_bottom_temp + solar, 
                 data = model_data)
summary(full_model)
```
## Variable Selection
### Stepwise AIC Backwards
```{r}
library(MASS)

# Stepwise AIC (backward)
step_model <- stepAIC(full_model, direction = "backward")

# View the final model summary
summary(step_model)
```

### Stepwise AIC Forwards (for the sake of verifying results)
```{r}
null_model <- lm(lake_mix_temp ~ 1, data = model_data)

step_forward <- stepAIC(null_model, 
                        scope = list(lower = ~1, 
                                     upper = ~ t2m + cloud + lake_bottom_temp + solar),
                        direction = "forward")
```

### Stepwise BIC model
```{r}
n <- nrow(model_data)
step_model_bic <- stepAIC(full_model, direction = "backward", k = log(n))
```

### Lasso Regression

```{r}
library(glmnet)

X <- as.matrix(model_data[, c("t2m", "cloud", "lake_bottom_temp", "solar")])
y <- model_data$lake_mix_temp

lasso_model <- cv.glmnet(X, y, alpha = 1)  # alpha=1 is lasso
plot(lasso_model)
coef(lasso_model, s = "lambda.min")  # coefficients at best lambda
coef(lasso_model, s = "lambda.1se")
```

Key finding: Lasso kept all 4 predictors — none were shrunk to zero. However, solar is extremely close to zero (-2.656e-07), suggesting it contributes very little. This aligns with it being the weakest predictor in the  AIC/BIC results.

**Variable Selection Results** 
Stepwise AIC (forwards and backwards) - Keep all 4 predictors
Stepwise BIC - Keep all 4 predictors
Lasso - Keep all 4 predictors


## Finding VIF 
```{r}
library(car)

# Calculate VIF
vif(full_model)
```

All VIFs are less than 2, which indicates that multicollinearity is not a problem in our model. 


## Testing for Interactions

### Plausible  Interactions
1. solar × cloud — most physically intuitive

Cloud cover directly blocks solar radiation, so the effect of solar on lake temperature likely depends on how cloudy it is
This is probably your strongest candidate for an interaction

2. t2m × cloud

Cloud cover affects air temperature (insulating effect), so their combined effect on lake mixed layer temp may not be purely additive

3. t2m × solar

Higher air temps combined with more solar radiation could have a compounding warming effect on the lake

### Full Interaction Model
```{r}
# Model with all two-way interactions
interaction_model <- lm(lake_mix_temp ~ (t2m + cloud + lake_bottom_temp + solar)^2,
                        data = model_data)
summary(interaction_model)

# Compare to full model without interactions
AIC(full_model, interaction_model)
anova(full_model, interaction_model)
```

Interactions t2m:lake_bottom_temp (p-value = 4.56e-07) and cloud:solar (p-value = 0.0453) were found to be significant at the 0.05 level. 

### Refined model with significant interactions

```{r}
refined_model <- lm(lake_mix_temp ~ t2m + cloud + lake_bottom_temp + solar +
                      t2m:lake_bottom_temp +
                      cloud:solar,
                    data = model_data)
summary(refined_model)
AIC(full_model, refined_model, interaction_model)
```

AIC is much lower than full model (84.94), and adjusted R^2 is now 0.7926. 

## Investigating Quadratic (Squared) Terms

### Scatter plots with linear and quadratic fits for all 4 predictors
```{r}
library(ggplot2)
library(car)

# 1. Scatter plots with linear and quadratic fits for all 4 predictors
p1 <- ggplot(model_data, aes(x = t2m, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = "Linear"), se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "t2m vs lake_mix_temp", color = "Fit") +
  theme_minimal()

p2 <- ggplot(model_data, aes(x = lake_bottom_temp, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = "Linear"), se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "lake_bottom_temp vs lake_mix_temp", color = "Fit") +
  theme_minimal()

p3 <- ggplot(model_data, aes(x = cloud, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = "Linear"), se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "cloud vs lake_mix_temp", color = "Fit") +
  theme_minimal()

p4 <- ggplot(model_data, aes(x = solar, y = lake_mix_temp)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "lm", formula = y ~ x, aes(color = "Linear"), se = FALSE) +
  geom_smooth(method = "lm", formula = y ~ poly(x, 2), aes(color = "Quadratic"), se = FALSE) +
  scale_color_manual(values = c("Linear" = "red", "Quadratic" = "blue")) +
  labs(title = "solar vs lake_mix_temp", color = "Fit") +
  theme_minimal()

# Print all 4 in a 2x2 grid
library(gridExtra)
grid.arrange(p1, p2, p3, p4, ncol = 2)

```

### Full quadratic model
```{r}
full_model_quad <- lm(lake_mix_temp ~ t2m + I(t2m^2) + 
                        cloud + I(cloud^2) + 
                        lake_bottom_temp + I(lake_bottom_temp^2) + 
                        solar + I(solar^2) + 
                        t2m:lake_bottom_temp + cloud:solar, data = model_data)
summary(full_model_quad)
```
The quadratic term I(t2m^2) is significant at the 0.05 level (p-value = 0.0012). 


### Refined quadratic model
```{r}
# Add quadratic terms
refined_model_quad <- lm(lake_mix_temp ~ t2m + I(t2m^2) + cloud + 
                           lake_bottom_temp + solar + t2m:lake_bottom_temp + cloud:solar, data = model_data)
summary(refined_model_quad)
AIC(refined_model_quad)
```

## Checking Assumptions
### Diagnostic Plots 
```{r}
par(mfrow = c(2,2))
plot(refined_model_quad)
```
The diagnostic plots show that normality and constant variance assumptions are not met. This can be seen by the heavy tails of the Q-Q plot and the 

### Shapiro-Wilk Normality Test
```{r}
shapiro.test(residuals(refined_model_quad))
```

### Breusch-Pagan Test
```{r}
library(lmtest)
# Homoscedasticity (constant variance)
bptest(refined_model_quad)
```

### Durbin-Watson Test for Autocorrelation
```{r}
# Autocorrelation (important since your data may be time-based)
dwtest(refined_model_quad)  # Durbin-Watson test
```


## Trying Transformations

### Try Anomaly transformation 
Anomaly transformation — very standard in climate science, removes seasonal mean:
```{r}
model_data$lake_mix_temp_anom <- model_data$lake_mix_temp - mean(model_data$lake_mix_temp)

model_data$t2m_anom <- model_data$t2m - mean(model_data$t2m)

model_data$lake_bottom_temp_anom <- model_data$lake_bottom_temp - mean(model_data$lake_bottom_temp)

model_data$cloud_anom <- model_data$cloud - mean(model_data$cloud)
model_data$solar_anom <- model_data$solar - mean(model_data$solar)
```

```{r}
# Refined quad model with anomalies
refined_quad_model_anom <- lm(lake_mix_temp_anom ~ t2m_anom + I(t2m_anom)^2 + cloud_anom + lake_bottom_temp_anom + solar_anom +
                           t2m_anom:lake_bottom_temp_anom + 
                           cloud_anom:solar_anom,
                         data = model_data)

summary(refined_quad_model_anom)
par(mfrow = c(2,2))
plot(refined_quad_model_anom)
```
Not much improvement in diagnostics, RSE is higher (0.30) and R^2 less than regular refined quad model, try different method of transformation.

### Try Box-Cox Transformation
```{r}
library(MASS)
bc <- boxcox(refined_model_quad)
# Find optimal lambda
optimal_lambda <- bc$x[which.max(bc$y)]
optimal_lambda

# Apply Box-Cox transformation
model_data$lake_mix_temp_bc <- (model_data$lake_mix_temp^optimal_lambda - 1) / optimal_lambda

# Refit model with transformed response
model_boxcox <- lm(lake_mix_temp_bc ~ t2m + I(t2m^2) + cloud + 
                     lake_bottom_temp + solar + t2m:lake_bottom_temp + cloud:solar, data = model_data)

summary(model_boxcox)
par(mfrow = c(2,2))
plot(model_boxcox)
```
Not much improvement in diagnostics, transforming would add some predictive power to the model but just lose interpretability. 


## Investigating Outliers

### Influence Plot
```{r}
# Create influence Plot
influencePlot(refined_model_quad)
```
### Investigate observation 118
```{r}
# Inspect observation 118
model_data[118, ]
```

Observation 118 has the maximum value for lake bottom temperature. 

### Remove observation 118 which appears to be an outlier
```{r}
final_model <- lm(lake_mix_temp ~ t2m + I(t2m^2) + cloud + 
                    lake_bottom_temp + solar + t2m:lake_bottom_temp + cloud:solar,
                  data = model_data[-118, ])
summary(final_model)
par(mfrow = c(2,2))
plot(final_model)
AIC(final_model)
```


# 4. Out of Sample Prediction Validation

## Prepare the data:
### Create yearly blocks by creating a new 'year' column that groups consecutive months together

```{r}
set.seed(8901)
model_data$time_index <- 1:173 #Re-establish the time index for each observation
model_data$year <- 1940 + (model_data$time_index - 1) %/% 2
```

### Verify groupings by 'year' are accurate for pairs of time indexes
```{r}
head(model_data$year) 
```
### Filter out your outlier
```{r}
cv_data <- model_data[-118, ]
```

## Run the Grouped 10-Fold Cross Validation:

### Create folds based on the 'year' column:

```{r}
library(caret)
#Create 10 folds based on the 'year' column so years aren't split
folds <- groupKFold(cv_data$year, k = 10)

```
### Run the 10-Fold CV

```{r}
# Pass these specific folds to trainControl
train_control <- trainControl(index = folds, method = "cv")

# Run your final model
cv_model <- train(
  lake_mix_temp ~ t2m + I(t2m^2) + cloud + lake_bottom_temp + solar + 
                  t2m:lake_bottom_temp + cloud:solar,
  data = cv_data,
  method = "lm",
  trControl = train_control
)

print(cv_model)
```

